---
title: "RIBBiTR Database Connection Setup"
author: "Cob Staines"
date: today
toc: true
format:
  html: 
    code-line-numbers: false
  pdf:
    code-line-numbers: false
    geometry: 
      - top=30mm
      - left=30mm
---

# Motivation

-   Connect to the RIBBiTR (or another remote) database with ease and repeatability
-   Manage login credentials for ease and security, so they won't be lost or shared with your code.

# Database Connection setup for RStudio

## Load packages

### Using "librarian"

"librarian" is a package and library management package in R which makes it easier to install, load, update and unload packages to meet dynamic environment needs. There are other ways to download, load, and maintain packages in R (e.g. "install.packages()" and "library()", but we recommend librarian for its simplicity and portability.

```{r, message=FALSE}
# install and load "librarian" package if not already required
if (!require(librarian)) {
  install.packages("librarian")
  library(librarian)
}
```

librarian downloads and loads packages using the "librarian::shelf" function. Below are the minimal recommended packages to establish a connection to the RIBBiTR database.

```{r, echo=TRUE, eval=TRUE}
# minimal packages for establishing RIBBiTR DB connection
shelf(tidyverse, RPostgres, DBI, usethis, RIBBiTR-BII/ribbitrrr)
```

## Defining your database connection parameters

### Access your local .Renviron file
Your .Renviron file a local file where you can save and reference your login credentials for easy use within R and RStudio, without risking losing them or potentially sharing them on accident when you share your code.

```{r, message=FALSE}
# open your local .Reniron file
usethis::edit_r_environ()
```

### Save connections parameters locally

Copy the following database connection parameters to your .Renviron file, substituting your login credentials (user & password).

```{r, eval=FALSE}
# RIBBiTR DB credentials
ribbitr.dbname = "ribbitr"
ribbitr.host = "ribbitr.c6p56tuocn5n.us-west-1.rds.amazonaws.com"
ribbitr.port = "5432"
ribbitr.user = "[YOUR-LOGIN-HERE]"
ribbitr.password = "[YOUR-PASSWORD-HERE]"
```

Save and close .Renviron, and restart RStudio.

## Establish database connection

Now, using the ribbitrrr:HopToDB() function, let's establish a connection!

```{r}
# establish database connection
dbcon = HopToDB("ribbitr")
```

HopToDB() returns a database connection object ("dbcon"). Keep track of this, you will need it to explore and pull data later.

## Begin using your connection!

Try out your connection by loading table metadata from the database

```{r}
mdt = tbl(dbcon, Id("public", "all_tables")) %>%
  collect()
head(mdt)
```

## Also try
- For those managing multiple database connections, the HopToDB() function allows you to store and fetch various sets of login credentials with a single keyword. Just substitute "ribbitr" in the example above with your own keyword!
- Your login credentials can also be accessed explicitly anytime using Sys.getenv("ribbitr.dbname"), etc. In most cases the HopToDB() function is all you need, however.
